---
title: "R syntax and the tidyverse"
author: "Pablo Gomez"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, message = F)
library(dplyr); library(ggplot2); library(ggpubr);library(Pabloverse)
```

# R basics

R is like a calculator, we can make mathematical operations, for example:

```{r basic operations}
2 + 2
```

R is a object-oriented programming language, this means that we create objects that contain information. In R you can achieve the same results using different approaches, for example, to create an object we just type a name for the object and assign it a value using the operators `=` or `<-`. We can make operations with objects of the same type, for example:

```{r storing values}
x = 2 # create a new object with the = operator
y <- 2 # create a new object with the <- operator
x + y # make a operation with the objects
```

## Vectors in R

You can store more than one value using vectors, to create a vector of numbers we use `c()`.
For example, we will store a sequence of numbers from 5 to 10 using 2 different approaches and then ask R if the objects are the same.  
**tip:** using the keys "alt" + "-" will automatically add the operator `<-`. Choosing which assign operator to use is a matter of preference, I personally think that is easier reading code with the operator `<-`, but a lot of people uses `=`.  

```{r vectors}
x <- c(5, 6, 7, 8, 9, 10) # create a sequence form 5 to 10
y = 5:10 # create the same sequence but with a different approach
x == y # ask R if the objects have the same information
```

When we have a vector, we can ask R specific values inside an object.

```{r vector operations}
# Here we ask the 3rd value from our sequence
x[3]
# Now we multiply the 3rd value of the x sequence times the 5th value of the y sequence
x[3] * y[5]
```

## Functions

R has a lot of base functions, but we can define new functions. When using R studio, the key *Tab* will help us to auto complete, this can help us a lot when we don't remember the exact name of the functions available.
The best part of programming with R is that it has a very active community. Since its open source, anyone can create functions and compile them in a package (or library). we can download these packages and access new functions.  
Functions in R require arguments, which we can see in the function documentation or if we press the key *Tab* when we are inside the function.  

```{r functions}
# To get the sum of a vector of numbers inside an object we use sum()
sum(x)
```

We can put functions inside function, for example, to get $\sqrt{\sum_1^n x}$ the square root of a sum of the numbers in x we can use:

```{r nested functions}
sqrt(sum(x))
```

Making functions in R is not as complicated as it sounds and can be very useful when we need to do repetitive work. To define a function we need to include the arguments that we want for the function and what are we doing with those arguments. For example, the following function has only one argument which is a name (string) and just pastes some text before and after:

```{r creating a function}
F1 <- function(name){
  paste("Hola", name, "! welcome to the R world!") # paste the name with some text
}

# Testing the function (Put your name)
F1(name = "Pablo")
```

Besides storing numbers in the objects in R, we can store text, matrices, tables, spatial objects, images, and other types of objects. Next we will import our own data and do some manipulation in R.

**Exercise:** Create a function that performs the $\sqrt{\sum_1^n x}$ operation you did previously with the code `sqrt(sum(x))`

```{r Excercise 1, eval=FALSE, include=FALSE}
# Define the function
F2 <- function(x){
  sqrt(sum(x))
}

# Test the function
F2(x)
```

## Types of variables

There are different types of variables in R, so far we have used the numeric and string types. If yo want to know what kind of variable is a given object, you can use the function `class()`. Lets try it.

```{r variables I}
class(x)
```

**Exercise: ** What kind of variable is the output from the first function you defined `F1()`?

```{r exercise vars, eval=FALSE, include=FALSE}
class(F1(name = 'Pablo'))
```

There are other type of variables that we will see later in the course. 

<!-- **Introduce different type of variables in R (e.g. Numeric, String, Factor, Date)** -->

## Introducing the pipes

The library `dplyr` has several functions that can help to clean, create new variables, and modify our data in other ways.  

```{r message = F}
# if we dont have installed the library we will need to install it using:
# install.packages("dplyr")
# we load the library:
library(dplyr)
```

`dplyr` introduces a new operator called *pipes* (`%>%`), which can connect several functions to an object. This is an alternative to write several functions in a single "line of code" in a more organized way. For example, if we want to execute a function `F1()` followed by another function `F2()` for the object `x`:  

`F2(F1(x))` is equivalent to `x %>% F1() %>% F2()`    

As you can notice, to read the code `F2(F1(x))` we have to go from the inside to the outside to see the order of execution of the functions, but when we read `x %>% F1() %>% F2()` we read from left to right, which is the same way we normally would read any text in western language.    

**Suggestion:** we can use the keys *Ctrl* + *shift* + *m* to insert the `%>%` operator.  

```{r pipes} 
# We previosly used this code to calculate the square root of the sum of x
sqrt(sum(x))
```

Using the pipes we can do the same more organized, by writing the order of execution from left to right. 

```{r pipes II}
x %>% # First we call the data
  sum() %>% # Sum all the values
  sqrt() # Compute the square root
```

You will notice that the outputs are exactly the same. Feel free to use whatever syntax you prefer, but for this course we will use mostly the pipes and writing the code from left to right.

# Importing datasets

R can import data in different formats. The most common are excel files (*.csv*, *.xls* y *.xlsx*), text files *.txt* and spatial data *.shp*, which we will talk about more in detail later.  
To read *.xls*, *.xlsx* and *.shp* files we will need to install some libraries. To install a new library you need to be connected to the internet and use the function `install.packages()` to install the library. Once it has been installed, you can load the library using the function `library()`.  
  
Download the excel file from this [link](https://ucdavis.app.box.com/folder/174771408063?s=cltv7bcwblxw5v90z5l56ognt9dprehv). It's not necessary to have a Box account.  
  
*Suggestion:* Sometimes when we want to use only one function from a library, we can just write the name of the library followed by the operator `::` and the name of the function, for example: `package::function()`. This way we won't have to load the whole library.  

```{r}
# If we dont have the library installed, we use:
# install.packages("readxl")
library(readxl) # load the library

# Import the excel file
PRRS <- readxl::read_xlsx("Data/PRRS.xlsx")
```

The most popular format for tables in R are called `data.frame`, when we import the data from a *.csv* o *.xlsx*. We can examine what kind of object it is using the function `class()`, an object can have more than one type of class.  

```{r}
class(PRRS)
```

# Data manipulation

In the following section we will cover some basics of data manipulation, this includes create, modify and summarise variables in our data. 

## Grouping and summarizing tables

```{r}
# We can count how many observations for each of a variable name
PRRS %>% 
  count(Result)
```

The previous table show us that were 986 observations with test result negative and 367 with test result positive. We can add multiple variables to count

```{r}
# We can count based on multiple variables
PRRS %>% 
  count(Result, Sex)
```

```{r}
PRRS %>% 
  count(Sex, wt = Result)
```



## Creating new variables

To create a new variable we can use the function `mutate()`. For example, lets create a new variable that will tell us if the farm type is a sow farm or not. For this we use the variable `farm_type` which already contains information for different farm types 

```{r}
PRRS %>% 
  mutate(SowFarm = ifelse(farm_type == 'sow farm', 1, 0))
```

## Joining datasets

```{r}
# Importing the farm dataseet
nodes <- read.csv("Data/node_attrib.csv")
# Importing the movement dataset
mov <- read.csv("Data/network.csv")
# Get the number of outgoing and incoming shipments 
Out <- mov %>% 
  group_by(id_orig) %>%
  summarise(Outgoing = n()) %>%
  rename(id = id_orig)
In <- mov %>%
  group_by(id_dest) %>%
  summarise(Incoming = n()) %>% 
  rename(id = id_dest)

# Joining the two datasets
nodes <- nodes %>% 
  left_join(Out, by ="id")

nodes <- nodes %>%
  left_join(In, by = "id")
nodes
```

Now we will get the ones that had the higher number of movements: 

```{r}
mov %>% 
  group_by(id_orig) %>% 
  summarise(N = n(), N_pigs = sum(pigs.moved)) %>%
  arrange(desc(N)) %>%
  head()
```
  

## Wide and Long formats

# Tables in R

We can use the base R functions to create basic tables. For example, we can create a table that shows the number of positive by sex:  

```{r}
TBL <- table(PRRS$Sex, PRRS$Result)
TBL
```

When we have a table object we can ask for a specific column or row, where is represented as [row, column].  

```{r}
TBL[,2] # Get the second column
```


------------------

[Go Back](index.html)