---
title: "R syntax and the tidyverse"
author: "Pablo Gomez"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, message = F)
library(dplyr); library(ggplot2); library(ggpubr);library(Pabloverse)
```

# R basics

R is like a calculator, we can make mathematical operations, for example:

```{r}
2 + 2
```

R is a object-oriented programming language, this means that we create objects that contain information. In R you can achieve the same results using different approaches, for example, to create an object we just type a name for the object and assign it a value using the operators `=` or `<-`. We can make operations with objects of the same type, for example:

```{r}
x = 2 # create a new object with the = operator
y <- 2 # create a new object with the <- operator
x + y # make a operation with the objects
```

## Vectors in R

You can store more than one value using vectors, to create a vector of numbers we use `c()`.
For example, we will store a sequence of numbers from 5 to 10 using 2 different approaches and then ask R if the objects are the same.  
**tip:** using the keys "alt" + "-" will automatically add the operator `<-`. Choosing which assign operator to use is a matter of preference, I personally think that is easier reading code with the operator `<-`, but a lot of people uses `=`.  

```{r}
x <- c(5, 6, 7, 8, 9, 10) # create a sequence form 5 to 10
y = 5:10 # create the same sequence but with a different approach
x == y # ask R if the objects have the same information
```

When we have a vector, we can ask R specific values inside an object.

```{r}
# Here we ask the 3rd value from our sequence
x[3]
# Now we multiply the 3rd value of the x sequence times the 5th value of the y sequence
x[3] * y[5]
```


## Introducing the pipes

The library `dplyr` has several functions that can help to clean, create new variables, and modify our data in other ways.  

```{r message = F}
# if we dont have installed the library we will need to install it using:
# install.packages("dplyr")
# we load the library:
library(dplyr)
```

`dplyr` introduces a new operator called *pipes* (`%>%`), which can connect several functions to an object. This is an alternative to write several functions in a single "line of code" in a more organized way. For example, if we want to execute a function `F1()` followed by another function `F2()` for the object `x`:  

`F2(F1(x))` is equivalent to `x %>% F1() %>% F2()`    

As you can notice, to read the code `F2(F1(x))` we have to go from the inside to the outside to see the order of execution of the functions, but when we read `x %>% F1() %>% F2()` we read from left to right, which is the same way we normally would read any text.  

**Suggestion:** we can use the keys *Ctrl* + *shift* + *m* to insert the `%>%` operator.  

## Types of variables

## Functions

R has a lot of base functions, but we can define new functions. When using R studio, the key *Tab* will help us to auto complete, this can help us a lot when we don't remember the exact name of the functions available.
The best part of programming with R is that it has a very active community. Since its open source, anyone can create functions and compile them in a package (or library). we can download these packages and access new functions.  
Functions in R require arguments, which we can see in the function documentation or if we press the key *Tab* when we are inside the function.  

```{r}
# To get the sum of a vector of numbers inside an object we use sum()
sum(x)
```

We can put functions inside function, for example, to get $\sqrt{\sum_1^n x}$ the square root of a sum of the numbers in x we can use:

```{r}
sqrt(sum(x))
```

Making functions in R is not as complicated as it sounds and can be very useful when we need to do repetitive work. To define a function we need to include the arguments that we want for the function and what are we doing with those arguments. For example, the following function has only one argument which is a name (string) and just pastes some text before and after:

```{r}
F1 <- function(name){
  x <- paste("Hola", name, "! welcome to the R world!") # paste the name with some text
  print(x)
}
# trying the function (Put your name)
F1(name = "Pablo")
```

Besides storing numbers in the objects in R, we can store text, matrices, tables, spatial objects, images, and other types of objects. Next we will import our own data and do some manipulation in R.


# Importing datasets

R can import data in different formats. The most common are excel files (*.csv*, *.xls* y *.xlsx*), text files *.txt* and spatial data *.shp*, which we will talk about more in detail later.  
To read *.xls*, *.xlsx* and *.shp* files we will need to install some libraries. To install a new library you need to be connected to the internet and use the function `install.packages()` to install the library. Once it has been installed, you can load the library using the function `library()`.  
  
Download the excel file from this [link](https://ucdavis.app.box.com/folder/174771408063?s=cltv7bcwblxw5v90z5l56ognt9dprehv). It's not necessary to have a Box account.  
  
**Suggestion:** Sometimes when we want to use only one function from a library, we can just write the name of the library followed by the operator `::` and the name of the function, for example: `package::function()`. This way we won't have to load the whole library.  

```{r}
# If we dont have the library installed, we use:
# install.packages("readxl")
library(readxl) # load the library

# Import the excel file
PRRS <- readxl::read_xlsx("Data/PRRS.xlsx")
```

The most popular format for tables in R are called `data.frame`, when we import the data from a *.csv* o *.xlsx*. We can examine what kind of object it is using the function `class()`, an object can have more than one type of class.  

```{r}
class(PRRS)
```


# Tables in R

We can use the base R functions to create basic tables. For example, we can create a table that shows the number of positive by sex:  

```{r}
TBL <- table(PRRS$Sex, PRRS$Result)
TBL
```

When we have a table object we can ask for a specific column or row, where is represented as [row, column].  

```{r}
TBL[,2] # Get the second column
```


# Data manipulation

## Creating new variables

## Joining datasets

## Wide and Long formats

## Grouping and summarizing tables

------------------

[Go Back](index.html)